Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-08-01 manuscript_english.tex
Edits to the Intro

Genetic variation among populations exhibits biogeographic patterns. In
this paper we establish a theoretical framework based on environmental
Grinellian niches as a tool for modeling species distributions, as well
as their projected past environmental scenarios to relate the present
evolutionary history in both the geographic distribution and genetic
structure of their populations. Thanks to the advance of next-generation
sequencing technologies, as well as more accurate climate models and an
enormous amount of information stored in biological collections, it is
possible to implement this theoretical framework with the help of the
joint analysis of spatial, environmental and genetic data, with tools
that are widely distributed among biogeographers and evolutionary
biologists. Our results reveal that when we have case of isolation by
distance, suitability prevalence along time corresponds to the fixation
index of populations with respect to a source population. Populations
closer to the historical endemic area present a higher genetic diversity
and a lower \(F_{st}\) value. This framework allows us to add a
biogeographic explanation to the results obtained with population
genetic methods and to generate maps of this structure as tools to
support conservation with a perspective that integrates both population
genetics and historical patterns of species distribution.

@@

The current distributions of species do not always correspond to their
historical distributions over evolutionarily significant periods. This
is because environmental conditions are not static over time; species
tend to distribute where conditions are most favorable.

It follows from this relationship between environment and species
distribution that population size also varies with time. This variation
in population size is related to the effective population size \(N_e\)
(Karlin 1968), so that an index reflecting changes in environmental
conditions in geography might be expected to be related to \(N_e\) and
thus to indicators of population structure such as the fixation index
\(F_{st}\). Thus, it is possible to relate patterns of changes in the
distribution of species to the genetic structure of their populations
using a statistical model that explains this relationship.

With this approach we can predict the geographic pattern of population
structure from environmental information. This approach is strongly
driven by advances in currently available climate simulations (Leonardi
et al. 2023; Krapp et al. 2021), as well as next generation sequencing
data, and supported by both ecological niche (Thorup et al. 2021;
Nogués-Bravo 2009) and population genetics theories (Lira-Noriega and
Manthey 2014).

In this work we propose a method to find the Suitability Prevalence Area
(SPA) as an index with a double purpose: 1) to find endemic areas to
delimit the historical distribution of the species and 2) to explain the
patterns of genetic diversity.

To obtain the SPA, we performed a historical reconstruction of the
geographical range back to 120 000 BC at 2 000 year intervals and
recorded the environmental suitability at each site.

@@

\hypertarget{fixation-index-as-a-functino-of-spa}{%
\subsubsection{Fixation index as a functino of
SPA}\label{fixation-index-as-a-functino-of-spa}}

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_st\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

\begin{longtable}[]{@{}ccc@{}}
\toprule\noalign{}
Population & R squared & p value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{S. a. aberti Carson-SFW} & 0 & 0.962 \\
\emph{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
\emph{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
\emph{S. a. aberti San Juan} & 0 & 0.952 \\
\emph{S. a. barberi} & 0.794 & 0.001 \\
\emph{S. a. chuscensis E} & 0.002 & 0.892 \\
\emph{S. a. chuscensis W} & 0.018 & 0.713 \\
\emph{S. a. ferreus Carson E} & 0.04 & 0.581 \\
\emph{S. a. ferreus Pike} & 0 & 0.967 \\
\emph{S. a. kaibabensis} & 0.005 & 0.841 \\
\end{longtable}

>>>>

Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established. In
this paper, we establish a theoretical framework based on projecting environmental
Grinellian niches back through time to relate the present geographic distribution of population genetic
structure to a given species' historical evolutionary context. Thanks to the advance of next-generation
sequencing technologies, as well as more accurate climate models and an
enormous amount of information stored in biological collections, it is
possible to implement this theoretical framework directly. We use the Tassel-earred Squirrel (\textit{Sciurus aberti}) as a model to jointly analyze spatial, environmental and genetic data and predict the historical endemic area of this species. Our results reveal that in cases of genetic isolation by
geographic distance, suitability prevalence along time corresponds to the genetic fixation
index of populations with respect to a source population. Populations
closer to the historical endemic area present a higher genetic diversity
and a lower \(F_{st}\) value. This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a
biogeographic explanation to the results obtained with population
genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to
support biodiversity conservation efforts. Overall, this work advances a perspective that integrates population
genetics with historical patterns of species distribution.

@@

Current species distributions do not always correspond to their
historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS). This
is because environmental conditions are highly dynamic over time, and species
tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (REF).

It follows from this relationship between environment and species
distribution that population size also varies with time. This variation
in population size is related to the effective population size \(N_e\)
(Karlin 1968), so that an index reflecting changes in environmental
conditions in geography might be expected to be related to \(N_e\) and
thus to indicators of population structure such as the fixation index
\(F_{st}\). It should therefore be possible to relate patterns of change in the
distribution of species to the genetic structure of their populations
using a statistical model that explains this relationship.

With this approach we expect to predict the geographic pattern of population-genetic
structure from abiotic environmental information alone. This approach is strongly
driven by advances in currently available climate simulations (Leonardi
et al. 2023; Krapp et al. 2021), as well as next-generation sequencing
data, and supported by both ecological niche (Thorup et al. 2021;
Nogués-Bravo 2009) and population genetics theories (Lira-Noriega and
Manthey 2014). Importantly, these predicted patterns of population structure will serve as a type of \textit{a priori} hypothesis for the influence of abiotic factors alone, with the empirical detection of deviations from this hypothesis indicating that biotic factors such as competition or predation from other species have also impacted the observed population structure.

In this work, we propose a method to find what we call the Suitability Prevalence Area
(SPA) that is the time-averaged area of highest environmental suitability for a given species. The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of the target species; and 2) predicting the expected patterns of genetic diversity given the inferred history of population stability and connectivity.

To obtain the SPA, we overlapped historical reconstructions of the
geographical distribution of a target species estimated every 2,000 years from the present day until 120,000 years ago, in each time interval recording the environmental suitability at each modeled site.

@@

\hypertarget{fixation-index-as-a-function-of-spa}{%
\subsubsection{Fixation index as a function of
SPA}\label{fixation-index-as-a-functino-of-spa}}

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_{st}\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, geografic references were
obtained for each population considered in the fixation index matrix.
The SPA value for each population was
extracted for each population site. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

\begin{longtable}[]{@{}ccc@{}}
\toprule\noalign{}
Population & R squared & p value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textit{S. a. aberti Carson-SFW} & 0 & 0.962 \\
\textit{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
\textit{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
\textit{S. a. aberti San Juan} & 0 & 0.952 \\
\textit{S. a. barberi} & 0.794 & 0.001 \\
\textit{S. a. chuscensis E} & 0.002 & 0.892 \\
\textit{S. a. chuscensis W} & 0.018 & 0.713 \\
\textit{S. a. ferreus Carson E} & 0.04 & 0.581 \\
\textit{S. a. ferreus Pike} & 0 & 0.967 \\
\textit{S. a. kaibabensis} & 0.005 & 0.841 \\
\end{longtable}
